Short and Mechanized Logical Relation for Dependent Types

### 2024-02-26 paper/paper.tex
pass through introduction. We still need to refine our contributions summary

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied to reason about dependent type
theories.
Tutorial material on this topic~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types, pierce2004advanced,harper2016practical} is primarily focussed on
systems with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.
\scw{need one sentence to motivate why logical relations are important}

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that proofs that use
logical relations are not applicable.
But this is not the case. Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code, these definitions are easy to miss and difficult to
isolate from their surrounding contexts. As a result, they are not accessible
to casual readers.

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set theory notation and terminology so that
the material is accessible to readers with general mathematical background,
That said, our explanations do not stray too far away from our proof scripts.
We label each lemma directly to their counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
Our typeset representation are purposefully made to directly follow our
mechanical proofs, while avoiding, as much as possible, the introduction
of artifacts that are specific to certain theorem prover.

@@

In Section~\ref{sec:spec}, we introduce \lang{}, the object language
of interest, and establish some elementary properties about parallel
reduction, which will be used as the equational theory of our language
and play a central role in defining our logical relation. The design
of \lang{} is reminiscent of Martin-Löf style predicative type
systems. The untyped conversion rule of \lang{} is inspired by Pure
Type Systems~\citet{barendregt1991introduction}, though we discuss how
our proof can be easily adjusted for systems with typed judgmental
equality.  We design \lang{} to incorporate some of the most common
features of dependent type theory, including an intensional identity
type, a boolean type, large elimination, dependent elimination, and a
full universe hierarchy. We believe a good coverage of features will
make our proof applicable to a broad range of type systems.

In Section~\ref{sec:logreldep}, we formulate the consistency property
and motivate why proof by logical relation is needed to derive it.
We define our logical relation as an inductively defined relation. We first
show that the logical relation respects evaluation (referred to as the
irrelevance property in \citet{martin-lof-a-la-coq}), then proceed to
prove that the logical relation is a partial function; that is, each
type can have at most one unique interpretation.

In Section~\ref{sec:logrelproof}, we give the definition of semantic
typing in terms of our logical relation so we can state and prove the
fundamental theorem, which states that syntactic typing implies
semantic typing, from which consistency follows as a corollary.

In Section~\ref{sec:extension}, we extend our logical relation to
prove the stronger property that every well-typed term has a
$\beta\eta$-normal form. After laying the foundation for handling
type-level computation in Section~\ref{sec:logreldep}, the extension
to include open terms in the logical relation requries mechanical
changes that directly mirrors the same extension for a simply typed
language. We use this extension to demonstrate the idea that once we
have established the base technique for proof by logical relation for
dependent types, we can factor out the complexities of an extension
the are not specifically related to dependent types; such an
extension, if desired, may be studied in the context of a simply typed
language and later ported into a dependently typed setting. We believe
it is possible to extend our logical relation to a Kripke-style and
relational model in a similar fashion.

@@

We hope our success at creating a short and mechanized proof for a
relatively feature-complete language will encourage future researchers
to leverage the tool of logical relation more often in mechanized
proofs for dependent types.
% % For
% % example, \citet{geuvers1994short} requires impredicativity to encode
% % Calculus of Constructions, an impredicative dependent type theory, but
% % also relies on the classification property to define the logical
% % relation over well-formed types and kinds. Their simple
% % presentation of the logical relation as an inductively defined
% % function over types, reminiscent of logical relations for simply typed
% % languages,

@@

\scw{The end of this section needs to include an explicit summar of the
contributions of the paper. }

@@

In this section, we present the dynamics and statics of the
dependent type theory whose logical consistency will be proven in
Section~\ref{sec:logrelproof}. For concision, we refer to this system
as \lang.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.
% \scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
% \yl{need at least 4 vars for J. removed the unnecessary t and C}
We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper since it is more amenable to
automation (Section~\ref{sec:automation}) and leaves no ambiguity
about variable freeness conditions.

>>>>

Our goal with this work is to demonstrate how the proof technique of
\emph{syntatic logical relations} can be applied to reason about dependent
type theories.  Logical relations are a powerful proof technique for reasoning
about lambda calculi and their operational semantics. They have been used to
show diverse properties such as strong normalization, contextual equivalence,
representation independence, noninterference, compiler correctness, and the
decidability of conversion algorithms. \scw{add citations? emphasize that we are working 
with syntactic logical relations.}

Tutorial material on logical relations~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types,
  pierce2004advanced,harper2016practical} is primarily focussed on systems
with simple or polymorphic types. In that context, syntactic logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation. However, neither of these techniques can be used to define a
logical relation in the context of a predicative dependent type theory, so a
young researcher might be excused for thinking that proofs that use logical
relations are not applicable for such languages.

But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs for dependent type theories}
Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code. As a result, their use of logical relations arguments
difficult to isolate from their surrounding contexts, and thus, not accessible
to casual readers.

Thus, our paper provides a gentle and accessible introduction to a powerful
technique for dependent type theories. To promote the use of machine-assisted
reasoning, our development is accompanied by a short mechanized proof script,
of less than 1000 lines of code, developed using the Coq/Rocq proof
assistant~\cite{coq}.

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type type system and the results
that we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

The result of our work is a development that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set theory notation and terminology so that
the material is accessible to readers with general mathematical background,
That said, our explanations do not stray too far away from our proof scripts.
We label each lemma directly to their counterpart in the proof script,
anticipating that readers may wish to reference and validate our formalism.
Our typeset representation are purposefully made to directly follow our
mechanical proofs, while avoiding, as much as possible, the introduction
of artifacts that are specific to theorem provers.

@@

The key contributions of each section of our paper are as follows.
\scw{I'm still not happy with this overview / contributions list. }

In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
interest, reminiscent of Martin-Löf style predicative type systems. A key
design choice that impacts our proofs is the use of an untyped conversion
rule, inspired by Pure Type Systems~\citet{barendregt1991introduction}, and
specified through parallel reduction to normal form~\citet{barendregt}.
Because we believe that a good coverage of features makes our proof applicable
to a broad range of type systems, \lang{} incorporates some of the most common
features of dependent type theory, including an intensional identity type, a
boolean type, large elimination, dependent elimination, and a full universe
hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
discuss how our proof can be adjusted for systems with a typed, judgmental
equality.

In Section~\ref{sec:logreldep}, we formulate the consistency property and
motivate where a proof by logical relations is required. The most important
contribution of this section is the definition of the logical relation as an
inductively defined relation and demonstrate how to reason about this
definition. We also show that the logical relation respects evaluation
(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
prove that the logical relation is a partial function; that is, each type can
have at most one unique interpretation.

In Section~\ref{sec:logrelproof}, shows that our definitions can be the
subject of well-established reasoning techniques for logical relations. In
particular, we define a notion of semantic typing in terms of our logical
relation and prove the fundamental theorem, which states that syntactic typing
implies semantic typing. Consistency follows from this result as a corollary.

In Section~\ref{sec:extension}, extends the definition of the logical relation
to include terms with free variables, and proves that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, this extension requries mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently-typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:eta} we show how to incorporate the function $\eta$-law
into the conversion rule for \lang{}.

@@

We hope our success at creating a short and mechanized proof for a
relatively feature-complete language will encourage future researchers
to leverage the tool of logical relations in mechanized
proofs for dependent types.
% % For
% % example, \citet{geuvers1994short} requires impredicativity to encode
% % Calculus of Constructions, an impredicative dependent type theory, but
% % also relies on the classification property to define the logical
% % relation over well-formed types and kinds. Their simple
% % presentation of the logical relation as an inductively defined
% % function over types, reminiscent of logical relations for simply typed
% % languages,

@@

@@

In this section, we present the dynamics and statics of the
dependent type theory whose logical consistency will be proven in
Section~\ref{sec:logrelproof}. For conciseness, we refer to this system
as \lang.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.
We use the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper since it is more amenable to
automation (Section~\ref{sec:automation}) and leaves no ambiguity
about variable freeness conditions.
